Classical Planning with Simulators: \\ Results on the Atari Video Games

### 2015-04-30 ijcai15-paper/paper.tex
camera ready ijcai

\begin{document}
% The file aaai.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%
% \pdfinfo{
% /Title (Classical Planning Algorithms on the Atari Video Games)
% /Author (Nir Lipovetzky, Miquel Ramirez, Hector Geffner)}

@@

 \author{Nir Lipovetzky \\
        {\small The University of Melbourne }\\
     {\small Melbourne, Australia}\\
     {\footnotesize\url{nir.lipovetzky@unimelb.edu.au}}
%% footnote doesn't look nice ; this looks better
% although can't use Nir's unimelb address that is too long :-)
% 
%    {\tiny\url{nirlipo@gmail.com}}
    \And
    Miquel Ramirez \\
    {\small Australian National University} \\
    {\small Canberra, Australia}\\
   {\footnotesize\url{miquel.ramirez@anu.edu.au}}
    \And
    Hector Geffner \\
    {\small ICREA \& Universitat Pompeu Fabra }\\
    {\small Barcelona, Spain }\\
    {\footnotesize\url{hector.geffner@upf.edu}}
 %%%% \thanks{firstname.lastname}
}

@@

\begin{abstract}
The Atari 2600 games supported in  the Arcade Learning Environment~\cite{bellemare:jair2013}
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however,
cannot be used  off-the-shelf as there is no  compact PDDL-model of  the games, and action effects and goals are  not
known \emph{a priori}. Indeed, there are no explicit goals, and the planner must select actions on-line while interacting 
with a simulator that returns successor states and rewards. None of this precludes  the use of  blind lookahead algorithms 
for action selection like breadth-first search or  Dijkstra's yet such methods are not effective over large  state spaces. 
We thus turn to a different class of  planning methods  introduced recently  that  have been shown to be  effective for solving
large planning problems but which do not require  prior knowledge of state transitions and costs (rewards) or goals.
The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, \ldots,
where  $\IW(k)$ is  a breadth-first search in which a state is  pruned when it is not the first state in the  search  to make true 
a subset of $k$ atoms. The  empirical results over \numgames\ Atari games show that  $\IW(1)$ performs at the level of UCT, 
the state-of-the-art planning method in this domain, and suggest the potential of width-based methods for planning with
simulators when factored, compact action models are not available.
\end{abstract}

@@

While all  these games feature a known initial (RAM) state and actions that
have deterministic effects, the problem of selecting the next action
to be done cannot be addressed with off-the-shelf classical planners 
\cite{planning:book,geffner:book}. This is because there is  no compact PDDL-like encoding of the domain and 
the goal to be achieved in each game  is not given. Indeed, there are no goals but rewards, and 
the planner must select actions on-line while interacting with a simulator that just
returns successor states and rewards.
%\footnote{Actually, in the Atari games, the rewards $r(a,s)$ depend only
% on the state $f(a,s)$ that results from doing action $a$ in the state $s$.}

The action selection problem in the Atari games can be addressed  as a \emph{reinforcement learning} problem
\cite{sutton:book} over a deterministic MDP where the state transitions and rewards are not known, or alternatively,
as a  \emph{net-benefit planning problem} \cite{coles2012survey,keyder:jair09} with unknown state transitions  and  rewards. 
ALE supports the two settings: an \emph{on-line plannning setting} where  actions are selected
after a lookahead, and a \emph{learning setting}  that  must produce   controllers for mapping  states into
actions reactively without any lookahead. In this work we pursue the on-line planning approach where the lookahead
seeks  action sequences with  maximum  reward up to a certain horizon.

@@

The paper is organized as follows. We review the iterated width  algorithm and its properties,
look at the    variations of the algorithm that we used in the Atari games, present and analyze the  
experimental results, and discuss related and future work.

@@

\IW\ is a systematic and complete blind-search algorithm like breadth-first search (\BRFS) and iterative deepening (\ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the
search. This structured exploration  has proved to be very effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition.
Yet, this changes the definition of the domain.} For example, $37\%$ of the 37,921 problems
considered  in \cite{nir:ecai2012} are solved by $\IW(1)$ while $51.3\%$ are  solved by $\IW(2)$.
These are instances obtained from 37 benchmark domains by splitting problems with $N$ atomic
goals into $N$ problems with one atomic goal each.  Since $\IW(k)$ runs in time that is exponential in $k$, 
this means that almost $90\%$ of the 37,921 instances are  solved in time that is either linear or quadratic in the number of problem variables, 
which in such encodings are all \emph{boolean}. Furthermore, when the performance of \IW\   is compared with  
a  Greedy Best First Search guided by the additive heuristic $h_{add}$, it turns out that ``blind'' \IW\ solves as many problems as the informed 
search, 34,627 vs. 34,849, far ahead of  other blind search  algorithms like \BRFS\ and \ID\  that solve 9,010 and 8,762 problems each. 
Moreover, \IW\ is  faster and results in shorter plans than in the heuristic search.

@@

The number of nodes \emph{generated} by $\IW(1)$ is $n \times D \times b$ in the worst case,
where $n$ is the number of problem variables, $D$ is the size of their domains, and $b$ is the number of
actions per  state.  This same number in a  breadth-first search is not linear in $n$
but exponential. For the Atari games,  $n=128$, $D=256$, and $b=18$, so that the product is
equal to $589,824$, which is large but feasible. On the other hand, the number
of nodes generated by $\IW(2)$ in the worst case is  $(n \times D)^2 \times b$, which is
equal to $19,327,352,832$ which is too large, forcing us to consider only a tiny fraction of
such states. For classical planning problems, the growth in the number of nodes from
$\IW(1)$ to $\IW(2)$ is not that large, as the variables are boolean. Indeed, we could have
taken the state vector for the Atari games as a vector of 1024 boolean variables,
and apply these algorithms to that representation. The results would be different.
In such a case, $\IW(1)$ and $\IW(2)$  would generate $n' \times D' \times b$ and
$(n' \times D')^2 \times b$ states, which for $n'=1024$ and $D=2$
represent  $36,864$ and $75,497,472$ states respectively.
These are feasible numbers, but we haven't used  this representation 
under the assumption that the correlations among the bits in each one of
the 128 words of the state vectors are meaningful.
In summary, from the basic \IW\ algorithm we are testing only the first
call $\IW(1)$.

@@

We tested \IW(1) and \BFS\ over \numgames\  of the $55$ different games 
considered in \cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
We left out  \textsc{Skiing} as the reported figures apparently use a  different reward structure.}
%% check
% \footnote{
% Entries marked as  ``--'' in the table are   still running at the time of submission. 
% Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
% Eventually all of  the  $55$ games will  be included in the evaluation.}
The two  algorithms were used to play the  games in the \emph{on-line planning
setting} supported by ALE where we will compare them with the planning algorithms 
considered by BNVB; namely, breadth-first search and \UCT. ALE supports also a \emph{learning setting} 
where the goal is to learn controllers that map  states into actions without
doing any lookahead. Algorithms across  the two settings are thus not directly comparable
as they compute different things. Learning controllers appears as a more challenging problem
and it is thus not surprising that planning algorithms like \UCT\ tend to achieve a higher score than 
learning algorithms. In addition, the learning algorithms reported by BNVB tend to use 
the state of the screen pixels, while the planning algorithms,  use the state
of the RAM memory. It is not clear however whether one input representation is
more challenging than the other; for the learning algorithms, BNVB mention
that the results tend to be better for the screen inputs. Experiments were
run on a high-performance computation cluster run by the University of Melbourne, where each computing node consists of a $6$-core Intel Xeon 
E5-2440, with $2.4$ GHz clock speed, with 64 GBytes of RAM installed.

@@

Last, in the lookahead, \IW(1) and \BFS\ select an action every $5$ frames, while \UCT\  selects 
an action every frame.  This means that in order to explore a branch 
$300$ frames deep, UCT gets to choose $300$ actions, while \IW($1$) and \BFS\ 
get to choose $60$ actions, both however using the same $300$ frames from the
budget. For this, we  followed the setup  of \BRFS\ in BNVB 
that also selects actions every $5$ frames, matching the behavior
of the emulator that requests an action also every $5$ frames. Since the 
lookahead budget is given by a maximum number of (new) frames, and the time 
is mostly taken by calls to the emulator, this may not be the best choice  
for $\IW(1)$ and \BFS\ that may therefore not be exploiting all the 
options afforded by the budget. Interestingly, when \UCT\ is limited to one
action every $5$ frames, its performance is reduced by up to a $50$\% in
games where it performs very well (\CrazyClimber), and does not improve in those games where it performs very poorly (\Freeway).

@@

\noindent \textbf{Acknowledgments.} We thank the people involved in
the creation and maintenance of the Arcade Learning Environment that
have made this work possible. The research by Nir Lipovetzky is partially
funded by the Australian Research Council Linkage grant LP11010015.
}

>>>>

\begin{document}

\pdfinfo{
/Title (Classical Planning with Simulators: Results on the Atari Video Games)
/Author (Nir Lipovetzky, Miquel Ramirez, Hector Geffner)}

@@

 \author{Nir Lipovetzky \\
        { The University of Melbourne }\\
     { Melbourne, Australia}\\
     {\footnotesize\url{nir.lipovetzky@unimelb.edu.au}}
%% footnote doesn't look nice ; this looks better
% although can't use Nir's unimelb address that is too long :-)
% 
%    {\tiny\url{nirlipo@gmail.com}}
    \And
    Miquel Ramirez \\
    { Australian National University} \\
    { Canberra, Australia}\\
   {\footnotesize\url{miquel.ramirez@anu.edu.au}}
    \And
    Hector Geffner \\
    { ICREA \& U. Pompeu Fabra }\\
    { Barcelona, Spain }\\
    {\footnotesize\url{hector.geffner@upf.edu}}
 %%%% \thanks{firstname.lastname}
}

@@

\begin{abstract}
The Atari 2600 games supported in the Arcade Learning
Environment~\cite{bellemare:jair2013} all feature a known initial
(RAM) state and actions that have deterministic effects. Classical
planners, however, cannot be used off-the-shelf as there is no compact
PDDL-model of the games, and action effects and goals are not known
\emph{a priori}. Indeed, there are no explicit goals, and the planner
must select actions on-line while interacting with a simulator that
returns successor states and rewards. None of this precludes the use
of blind lookahead algorithms for action selection like breadth-first
search or Dijkstra's yet such methods are not effective over large
state spaces.  We thus turn to a different class of
planning methods introduced recently that have been shown to be
effective for solving large planning problems but which do not require
prior knowledge of state transitions, costs (rewards) or goals. The
empirical results over \numgames\ Atari games show that the simplest
such algorithm performs at the level of UCT, the state-of-the-art
planning method in this domain, and suggest the potential of
width-based methods for planning with simulators when factored,
compact action models are not available.

\end{abstract}

@@

While all  these games feature a known initial (RAM) state and actions that
have deterministic effects, the problem of selecting the next action
to be done cannot be addressed with off-the-shelf classical planners 
\cite{planning:book,geffner:book}. This is because there is  no compact PDDL-like encoding of the domain and 
the goal to be achieved in each game  is not given, precluding the automatic derivation of heuristic functions and other inferences. Indeed, there are no goals but rewards, and 
the planner must select actions on-line while interacting with a simulator that just
returns successor states and rewards.
%\footnote{Actually, in the Atari games, the rewards $r(a,s)$ depend only
% on the state $f(a,s)$ that results from doing action $a$ in the state $s$.}

The action selection problem in the Atari games can be addressed  as a \emph{reinforcement learning} problem
\cite{sutton:book} over a deterministic MDP where the state transitions and rewards are not known, or alternatively,
as a  \emph{net-benefit planning problem} \cite{coles2012survey,keyder:jair09} with unknown state transitions  and  rewards. 
ALE supports the two settings: an \emph{on-line plannning setting} where  actions are selected
after a lookahead, and a \emph{learning setting}  that  must produce   controllers for mapping  states into
actions reactively without any lookahead. In this work, we are interested in the on-line planning setting.

@@

The paper is organized as follows. We review the iterated width  algorithm and its properties,
look at the    variations of the algorithm that we used in the Atari games, and present the experimental results.

@@

\IW\ is a systematic and complete blind-search algorithm like breadth-first search (\BRFS) and iterative deepening (\ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the
search. This structured exploration  has proved to be very effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition.
Yet, this mapping changes the definition of the domain.} For example, $37\%$ of the 37,921 problems
considered  in \cite{nir:ecai2012} are solved by $\IW(1)$ while $51.3\%$ are  solved by $\IW(2)$.
These are instances obtained from 37 benchmark domains by splitting problems with $N$ atomic
goals into $N$ problems with one atomic goal each.  Since $\IW(k)$ runs in time that is exponential in $k$, 
this means that almost $90\%$ of the 37,921 instances are  solved in time that is either linear or quadratic in the number of problem variables, 
which in such encodings are all \emph{boolean}. Furthermore, when the performance of \IW\   is compared with  
a  Greedy Best First Search guided by the additive heuristic $h_{add}$, it turns out that ``blind'' \IW\ solves as many problems as the informed 
search, 34,627 vs. 34,849, far ahead of  other blind search  algorithms like \BRFS\ and \ID\  that solve 9,010 and 8,762 problems each. 
Moreover, \IW\ is  faster and results in shorter plans than in the heuristic search.

@@

The number of nodes \emph{generated} by $\IW(1)$ is $n \times D \times
b$ in the worst case, where $n$ is the number of problem variables,
$D$ is the size of their domains, and $b$ is the number of actions per
state.  This same number in a breadth-first search is not linear in
$n$ but exponential. For the Atari games, $n=128$, $D=256$, and
$b=18$, so that the product is equal to $589,824$, which is large but
feasible. On the other hand, the number of nodes generated by $\IW(2)$
in the worst case is $(n \times D)^2 \times b$, which is equal to
$19,327,352,832$ which is too large, forcing us to consider only a
tiny fraction of such states. For classical planning problems, the
growth in the number of nodes from $\IW(1)$ to $\IW(2)$ is not that
large, as the variables are boolean. Indeed, we could have taken the
state vector for the Atari games as a vector of 1024 boolean
variables, and apply these algorithms to that representation. The
number of atoms would indeed be much smaller, and both IW(1) and IW(2)
would run faster then. However by ignoring the correlations among bits
in each one of the 128 words, the results would be weaker.

@@

We tested \IW(1) and \BFS\ over \numgames\  of the $55$ different games 
considered in \cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
We left out  \textsc{Skiing} as the reported figures apparently use a  different reward structure.}
%% check
% \footnote{
% Entries marked as  ``--'' in the table are   still running at the time of submission. 
% Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
% Eventually all of  the  $55$ games will  be included in the evaluation.}
The two  algorithms were used to play the  games in the \emph{on-line planning
setting} supported by ALE where we will compare them with the planning algorithms 
considered by BNVB; namely, breadth-first search and \UCT. ALE supports also a \emph{learning setting} 
where the goal is to learn controllers that map  states into actions without
doing any lookahead. Algorithms across  the two settings are thus not directly comparable
as they compute different things. Learning controllers appears as a more challenging problem
and it is thus not surprising that planning algorithms like \UCT\ tend to achieve a higher score than 
learning algorithms. In addition, the learning algorithms reported by BNVB tend to use 
the state of the screen pixels, while the planning algorithms,  use the state
of the RAM memory. It is not clear however whether the use of one input representation is
more challenging than the use of the other. For the learning algorithms, BNVB mention
that the results tend to be better for the screen inputs. Experiments were
run on a cluster, where each computing node consists of a $6$-core Intel Xeon 
E5-2440, with $2.4$ GHz clock speed, with 64 GBytes of RAM installed.

@@

Last, in the lookahead, \IW(1) and \BFS\ select an action every $5$
frames, while \UCT\ selects an action every frame.  This means that in
order to explore a branch $300$ frames deep, UCT gets to choose $300$
actions, while \IW($1$) and \BFS\ get to choose $60$ actions, both
however using the same $300$ frames from the budget. For this, we
followed the setup of \BRFS\ in BNVB that also selects actions every
$5$ frames, matching the behavior of the emulator that requests an
action also every $5$ frames. Since the lookahead budget is given by a
maximum number of (new) frames, and the time is mostly taken by calls
to the emulator, this may not be the best choice for $\IW(1)$ and
\BFS\ that may therefore not be exploiting all the options afforded by
the budget. Interestingly, when \UCT\ is limited to one action every
$5$ frames, its performance is reduced by up to a $50$\% in games
where it performs very well (\CrazyClimber), and does not appear to
improve in those games where it performs very poorly (\Freeway).

@@

\noindent }

\section*{Acknowledgments} We thank the creators of the ALE. The research by Nir Lipovetzky is partially
funded by the Australian Research Council Linkage grant LP11010015.
